---
title: "In-class Exercise 4B: Spatial Interaction"
author: "Goh Si Hui"
date: 2023/12/09
date-format: long
date-modified: "last-modified"
format: html 
execute: 
  echo: true
  eval: true
  warning: false
editor: visual 
---

```{r}
pacman::p_load(tidyverse, sf, tmap, httr)
```

```{r}
flow_data <- read_rds("data/rds/flow_data_tidy.rds")
glimpse(flow_data)
```

rename morning peak to trips

rename dist to DIST because R has a function called dist

add in two columns to find out intrazonal and interzonal trips

intrazonal trips means that the push and pull factors are the same

spatial interaction should focus on interzonal. so we should filter out the intrazonal equals to 0

clear the env before doing the models!

dist is the impedience

add a '-1' to remove the intercept because origin/dest constrained dun need an intercept.

look at the impt of sch_count, biz_count, log(dist)

check the parameters alpha, beta and gamma values (see slide 8 of lect 3, gravity models)

-   distance must be negative, because rs is inverse. ppl dun wanna travel too far. if it is positive, then need to investigate further

-   log sch and log biz should always be positive

-   last col is the pvalues. make sure it is less than 0.05 so that we can accept these factors. If more, then need to say there are not statistically significant

use R square "goodness of fit" to explain how well the factors can explain the model

-   create a function with observed results and estimated results.

-   use correlation coef funct in base 2

-   then square the corr result to get r2.

-   helps to explain how well it explain the rate of flow,

rmse on how good it is at estimation. the results is the number of errors (check documentation of performance_rmse again!) chose to normalised to be false,so it will use the raw value (actual root mean square error)

Origin Constrained SIM

doubly constrained -\> don't have the attractiveness factors, don't need to minus 1 the intercept

smaller RMSE is better.

plot rmse, check the outlier and might want to further investigate by removing the outlier and run the model and check how it affects the rmse.
