---
title: "Hands-on Exercise 1A: Geospatial Data Wrangling with R"
date: 2023/11/13
date-format: long
---

## Overview

In this hands-on exercise, I will be sharing how to import and wrangle geospatial data using appropriate R packages.

## Getting Started

### Packages 

The code chunk below installs and loads sf and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse)

```

### Data Acquisition

The datasets we are using are downloaded from the following sources:

## Importing Geospatial Data

### Importing polygon feature data

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")

```

### Importing polyline feature data in shapefile form

The code chunk below uses st_read() function of sf package to import CyclingPath shape file into R as line feature data frame.

```{r}
cyclingpath = st_read(dsn = "data/geospatial", layer = "CyclingPathGazette")
```

### Importing GIS data in kml format 

We will also use st_read to import GIS data in kml format:

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.kml")
```
