---
title: "Hands-on Exercise 1B: Choropleth Mapping with R"
author: "Goh Si Hui"
date: 2023/11/16
date-format: long
date-modified: "last-modified"
format: html 
execute: 
  echo: true
  eval: true
  warning: false
editor: visual 
---

## Overview

In this hands-on exercise, I will be sharing how to plot functional and truthful [choropleth maps](https://alg.manifoldapp.org/read/introduction-to-cartography/section/c3c06272-8b8b-49e7-a957-da0d06550b73 "What is a Choropleth map?") using an R package [`tmap`](https://r-tmap.github.io/tmap/ "tmap: thematic maps in R").

## Getting Started

### Packages

For this exercise, we will be using `tmap`, [`sf`](https://r-spatial.github.io/sf/ "Package sf: Simple Features for R") and [`tidyverse`](https://www.tidyverse.org/ "Tidyverse: collection of R packages designed for data science") packages. The code chunk below loads `tmap`, `sf` and `tidyverse` packages into R environment.

```{r}
pacman::p_load(tmap, sf, tidyverse)

```

::: callout-note
Note that we will be using the readr, tidyr and dplyr packages which are part of the tidyverse package.
:::

### Data Acquisition

The datasets we are using are downloaded from the following sources:

-   Master Plan 2014 Subzone Boundary (Web): [data.gov.sg](https://beta.data.gov.sg/ "data.gov.sg").

-   Singapore Residents by Planning Area/Subzone, Age, Group, Sex and Type of Dwelling, June 2011 - 2020: [Department of Statistics, Singapore.](https://www.singstat.gov.sg "Department of Statistics")

## Importing Data

### Importing Geospatial Data

The code chunk below uses st_read() function of sf package to import the planning subzones, which is a polygon feature data frame.

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

Let's take a look at the contents of `mpsz` using the following code chunk:

```{r}
mpsz
```

### Importing Attribute Data

```{r}
popdata <- read_csv("data/aspatial/respopagesexfa2011to2020.csv")
```

Let's take a look at the data imported in:

```{r}
list(popdata)
```

Note: We can do some exploratory data analysis after importing the data. For example, plot charts to understand the population data.

## Data Preparation

Before a thematic map can be prepared, we need to prepare a data table with 2020 values. The data table should include the following variables:

-   PA

-   SZ

-   YOUNG: age group 0 to 4 until age group 20 to 24

-   ECONOMY ACTIVE: age group 20 to 29 up till age group 60 to 64

-   AGED: age group 65 and above

-   TOTAL: all age group

-   DEPENDENCY: the ratio between YOUNG and AGED against ECONOMY ACTIVE group

### Data Wrangling

We will be using the following functions from tidyverse package:

-   pivot_wider() of tidyr package, and

-   mutate(), filter(), group_by() and select() of dplyr package

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG,
              values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         + rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
         rowSums(.[13:15])) %>%
mutate(`AGED` = rowSums(.[16:21])) %>%
mutate(`TOTAL` = rowSums(.[3:21])) %>%
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)/ `ECONOMY ACTIVE`) %>%
select(`PA`,`SZ`, `YOUNG`,
        `ECONOMY ACTIVE`, `AGED`,
        `TOTAL`,`DEPENDENCY`)
```

### Joining Attribute Data and Geospatial Data

We will convert the values in PA and SZ files to uppercase because the values of PA and SZ fields are made up of upper and lower case while the SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ),
            .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Then, we will use left_join() of dplyr to join the geographical data and trribute table using planning subzone name (SUBZON_N and SZ) as the common identifiers.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## Choropleth Mapping Geospatial Data Using *tmap*

To prepare thematic map using tmap, we can:

-   Plot a thematic map quickly using qtm().

-   Plot highly customisable thematic map by using tmap elements.

### Plotting a choropleth map quickly using *qtm()*

The code chunk below draws a cartographic standard choropleth map.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

### Creating a choropleth map by using *tmap*'s elements

To draw a high quality cartographic choropleth map as shown in the figure below, tmap's drawing elements should be used.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          style = "quantile",
          palette = "Blues",
          title = "Dependency ratio") + 
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) + 
  tm_borders(alpha = 0.5) +
  tm_compass(type = "8star", size = 2) + 
  tm_scale_bar() +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Subzone boundary from Urban Redevelopment Authority (URA) \n and Population data from Department of Statistics (DOS)",
             position = c("left", "bottom"))

```

In the following sub-section, we will share with you tmap functions that used to plot these elements.

#### Drawing a base map

The basic building block of tmap is tm_shape() followed by one or more layer elemments such as tm_fill() and tm_polygons().

In the code chunk below, tm_shape() is used to define the input data (i.e mpsz_pop2020) and tm_polygons() is used to draw the planning subzone polygons.

```{r}
 tm_shape(mpsz_pop2020) + 
   tm_polygons()
```

#### Drawing a choropleth map using tm_polygons()

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, we can assign the target variable (e.g. Dependency) to tm_polygons().

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY")
```

#### Drawing a choropleth map using tm_fill() and tm_border()

tm_polygons is a wraper of tm_fill() and tm_border(). tm_fill() shades the polygons using the default colour scheme and tm_borders() adds the borders of the shapefile onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY") + 
  tm_borders(lw = 0.1, alpha = 1)
```

### Data Classification Methods of tmap

Most choropleth maps employ some methods of data classification. The point of classification is to take a large number of observations and group them into data ranges or classes.

tmap provides a total ten data classification methods, namely: fixed, sd, equal, pretty (default), quantile, kmeans, hclust, bclust, fisher, and jenks.

To define a data classification method, the style argument of tm_fill() or tm_polygons() will be used.

#### Plotting choropleth maps with built-in classification methods

The code chunk below shows a jenks data classification that used 5 classes

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          n = 5, 
          style = "jenks") + 
  tm_borders(alpha = 0.5)
```

The code chunk below uses the equal data classification method.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### Plotting choropleth map with custom break

We can also compute our own category breaks.

First, we will compute and display the descriptive statistics of DEPENDENCY field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Using the above results, we set the break points at 0.60, 0.70, 0.80, 0.90, and 0 will be the minimum while 1 will be the maximum.

We plot the choropleth map using the following code chunk:

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY", 
          breaks = c(0.0,0.60,0.70,0.80, 0.90, 1.00)) + 
  tm_borders(alpha = 0.5)
```

### Colour Scheme

tmap supports colour ramps either defined by the user or a set of predefined colour ramps from the RColorBrewer package.

#### Using ColourBrewer palette

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          n = 6,
          style = "quantile", 
          palette = "Blues") +
  tm_borders(alpha = 0.5)


```

To revese the colour shading, add a "-" prefix under "palette".

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          n = 6, 
          style = "quantile",
          palette = "-Greens") + 
  tm_borders(alpha = 0.5)
```

### Map Layouts

#### Map Legend

In tmap, several legend options are provided to change the placement, format and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          style = "jenks",
          palette = "Blues", 
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n (Jenks Classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45,
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) + 
  tm_borders(alpha = 0.5)
```

#### Map Style

tmap allows a wide variety of layout settings to be changed. They can be called by using tmap_style().

The code chunk below shows the classic style is used.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          style = "quantile",
          palette = "-Greens") + 
  tm_borders(alpha = 0.5) + 
  tmap_style("classic")
```

#### Cartographic Furniture

tmap also also provides arguments to draw other map furniture such as compass, scale bar and grid lines.

In the code chunk below, tm_compass(), tm_scale_bar() and tm_grid() are used to add compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", style = "quantile",
          palette = "Blues", 
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n (Based on Quantile)",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45,
            legend.width = 0.35,
            frame = TRUE) + 
  tm_borders(alpha = 0.5) + 
  tm_compass(type = "8star", size = 2) + 
  tm_scale_bar(width = 0.15) + 
  tm_grid(lwd = 0.1, alpha = 0.2) + 
  tm_credits("Source: Planning Subzone boundary from Urban Redevelopment Authority (URA) \n and Population data from Department of Statistics (DOS)",
             position = c("left", "bottom"))
```

```{r}
tmap_style("white")
```

### Drawing Multiple Small Choropleth Maps

In tmap, small multiple maps (also known as facet maps) can be plotted in three ways:

-   by assigning multiple values to at least one of the asthetic arguments,

-   by defining a group-by variable in tm_facets(), and

-   by creating multiple stand-alone maps with tmap_arrange().

#### By assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues")+
  tm_layout(legend.position = c("right", "bottom")) + 
  tm_borders(alpha = 0.5) + 
  tmap_style("white")
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(c("DEPENDENCY", "AGED"),
              style = c("equal", "quantile"),
              palette = list("Blues", "Greens")) + 
  tm_layout(legend.position = c("right", "bottom"))
```

#### By defining a group-by variable in tm_facets()

```{r}
tm_shape(mpsz_pop2020) + 
  tm_fill("DEPENDENCY", 
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords = TRUE,
            drop.shapes = TRUE) + 
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) + 
  tm_borders(alpha = 0.5)
```

#### By creating multiple stand-alone maps with tmap_arrange()

```{r}
youngmap <- tm_shape(mpsz_pop2020) + 
  tm_polygons("YOUNG", 
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020) + 
  tm_polygons("AGED", 
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

### Mapping Spatial Object Meeting a Selection Criterion

Instead of creating small multiple choropleth map, you can also use selection funtion to map spatial objects meeting the selection criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_fill("DEPENDENCY", 
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1)+
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"), 
            frame = FALSE) + 
  tm_borders(alpha = 0.5)
```
